---
title: "OptiDesign: Advanced Experimental Design for Breeding Trials"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OptiDesign: Advanced Experimental Design for Breeding Trials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# 1. Introduction

Experimental design is a fundamental component of plant breeding and agronomic research. The ability to accurately estimate genetic effects, compare treatments, and predict performance depends critically on how field trials are constructed.

Traditional designs such as randomized complete block designs (RCBD) and alpha-lattice designs often assume:

- regular block structures
- balanced replication
- independence between spatial and genetic effects

However, modern breeding programs face increasingly complex constraints:

- large numbers of genotypes
- limited field capacity
- heterogeneous environments
- availability of genomic and pedigree information

`OptiDesign` addresses these challenges by integrating:

- flexible field layout construction
- genetic structure (family, pedigree, genomic)
- spatial-aware arrangement
- optional mixed-model efficiency evaluation

into a unified design framework.

# 2. Conceptual Framework

## 2.1 Design Objective

Let:

- $y$ be the vector of observed phenotypes
- $g$ be the genetic effects
- $e$ be the residual errors

We consider the mixed model:

$$
y = X\beta + Zg + e
$$

where:

- $X$ is the design matrix for fixed effects
- $Z$ is the incidence matrix for genotypes
- $g \sim N(0, \sigma_g^2 K)$
- $e \sim N(0, R)$

The design problem consists of choosing $X$ and $Z$ through field layout construction to optimize:

- precision of $\hat{\beta}$
- prediction accuracy of $\hat{g}$

## 2.2 Field as a Spatial Process

Field plots are spatially structured. Residuals are rarely independent because of:

- row effects
- column effects
- spatial autocorrelation
- soil gradients
- management effects

Supported residual structures include:

- "IID"
- "AR1"
- "AR1xAR1"

## 2.3 Genetic Structure in Design

Entries can be grouped based on:

- family labels
- pedigree relationship matrix ($A$)
- genomic relationship matrix ($G$ or GRM)

This allows the design to:

- reduce clustering of related genotypes
- improve diversity spread
- reduce confounding between spatial and genetic signals

# 3. Design Strategies

## 3.1 Block-Based Designs with Repeated Checks (`prep_famoptg()`)

### Principle

`prep_famoptg()` constructs a block-based design in which checks are repeated in every block, while non-check treatments are allocated according to their specified replication levels.

Depending on the input structure, the same function can generate:

- an augmented design
- a partially replicated (p-rep) design
- an RCBD-type design

Let:

- $C$ = number of checks
- $B$ = number of blocks
- $r_i$ = replication number of non-check treatment $i$

The basic structure is:

- checks are replicated in all blocks
- non-check treatments are distributed across blocks according to their replication levels
- each non-check treatment appears at most once per block

This last rule is important. For example, if a treatment has replication 2 and the number of blocks is at least 2, its two replicates are placed in two different blocks rather than repeated within the same block.

### Special cases

The same function can represent different design classes depending on the inputs:

- **Augmented design**: when `p_rep_treatments` is `NULL` or empty and test entries are unreplicated while checks appear in all blocks.
- **Partially replicated design**: when some non-check entries are replicated and others are unreplicated.
- **RCBD-type design**: when all non-check treatments have the same replication number greater than 1 and are spread across blocks. In addition, `n_blocks` is equal to the replication number of non-check treatments.

### Advantages

- efficient use of field space
- broader genetic coverage
- improved estimation via checks
- one framework for augmented, p-rep, and RCBD-type layouts

## 3.2 Stream-Based Alpha Designs (`alpha_rc_stream()`)

### Principle

The field is treated as a linear stream:

1. define traversal order
2. build a global stream
3. split the stream into replicates
4. split each replicate into blocks

### Key property

Unused plots are placed only at the end of the stream.

# 4. Example Dataset

```{r}
data("OptiDesign_example_data", package = "OptiDesign")
x <- OptiDesign_example_data
names(x)
```

# 5. Example 1 — Family-based block design with repeated checks

This example illustrates a family-based call to `prep_famoptg()`. Depending on the replication pattern supplied in the example inputs, the same function may behave as an augmented design constructor, a p-rep constructor, or an RCBD-type constructor.

```{r}
out_fam <- do.call(
  OptiDesign::prep_famoptg,
  c(
    x$OptiDesign_famoptg_example,
    x$OptiDesign_famoptg_args_family
  )
)

dim(out_fam$layout_matrix)
head(out_fam$field_book)
```

## 5.1 Interpretation

This example highlights the general repeated-check block framework:

- checks are present in every block
- non-check treatments are allocated across blocks according to their replication levels
- no non-check treatment can occur twice in the same block
- grouping is based on family labels in this example

# 6. Example 2 — Stream-based alpha design

```{r}
out_alpha <- do.call(
  OptiDesign::alpha_rc_stream,
  c(
    x$OptiDesign_alpha_example,
    x$OptiDesign_alpha_args_family
  )
)

dim(out_alpha$layout_matrix)
head(out_alpha$field_book)
```

## 6.1 Interpretation

This example shows a fixed-grid alpha row-column stream design where:

- the field is traversed as a single ordered stream
- replicate segments are contiguous in that stream
- checks are placed in every incomplete block
- unused cells, if any, occur only at the end of the stream

# 7. Advanced Example — GRM-based design

```{r}
out_grm <- do.call(
  OptiDesign::alpha_rc_stream,
  c(
    x$OptiDesign_alpha_example,
    x$OptiDesign_alpha_args_grm
  )
)

head(out_grm$field_book)
```

This example illustrates how a genomic relationship matrix can be used to derive grouping information that influences layout construction.

# 8. Dispersion Optimization

A useful design objective is to reduce local relatedness among nearby plots.

Informally, if $d(i,j)$ denotes the similarity between plots $i$ and $j$, one may seek to reduce:

$$
\sum_{(i,j)\in \mathcal{N}} d(i,j)
$$

where $\mathcal{N}$ is the set of neighboring plot pairs.

Potential effects include:

- reduced local clustering
- improved robustness to spatial confounding
- better distribution of related materials across the field

# 9. Efficiency Evaluation

## Fixed effects

Under a fixed-effect perspective, precision depends on quantities such as:

$$
Var(\hat{\beta}) = (X^T V^{-1} X)^{-1}
$$

## Random effects

Under a random-effect perspective, a key quantity is the prediction error variance:

$$
PEV = Var(\hat{g} - g)
$$

Lower values indicate better precision or better prediction, depending on the model objective.

# 10. Practical Guidelines

## Use `prep_famoptg()` when:

- checks should appear in every block
- many entries must be screened
- replication resources are limited
- the design may be augmented, partially replicated, or RCBD-type depending on the treatment replication structure
- early-stage breeding evaluation is a main objective

## Use `alpha_rc_stream()` when:

- field dimensions are fixed
- operational field-book order matters
- contiguous replicate segments are desired
- incomplete blocks may be unequal
- unused cells should be pushed to the end of the stream

## Grouping choice

- Family → simple and interpretable structure
- GRM → genomic precision
- A matrix → pedigree-based structure

# 11. Summary

`OptiDesign` enables:

- realistic field design
- genetic-aware layout
- spatial optimization
- efficiency evaluation

The function `prep_famoptg()` should be understood as a general block-based repeated-check design constructor that can generate augmented, partially replicated, or RCBD-type layouts depending on the supplied replication pattern.

The function `alpha_rc_stream()` provides a distinct stream-based strategy for fixed-grid alpha row-column layouts under operational field constraints.

Together, these tools are especially useful in breeding programs where field implementation constraints and genetic structure should be considered jointly during design construction.

# 12. Session Information

```{r}
sessionInfo()
```