---
title: "OptiDesign: Advanced Experimental Design for Breeding Trials"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{OptiDesign: Advanced Experimental Design for Breeding Trials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Introduction

Experimental design is a fundamental component of plant breeding and agronomic
research. The ability to accurately estimate genetic effects, compare
treatments, and predict field performance depends critically on how trials are
constructed.

Traditional designs such as randomized complete block designs (RCBD) and
alpha-lattice designs typically assume regular block structures, balanced
replication, and independence between spatial and genetic effects. Modern
breeding programs face increasingly complex constraints that these classical
frameworks do not address well: large numbers of genotypes, limited field
capacity, heterogeneous environments, and the availability of genomic and
pedigree information.

`OptiDesign` addresses these challenges by integrating flexible field layout
construction, genetic structure (family, pedigree, genomic), spatial-aware
arrangement, and optional mixed-model efficiency evaluation into a unified
design framework.

---

## Conceptual Framework

### The mixed model

Let $y$ be the vector of observed phenotypes, $g$ the genetic effects, and
$e$ the residuals. The general mixed model underlying design evaluation is:

$$y = X\beta + Zg + e$$

where:

| Symbol | Description |
|--------|-------------|
| $X$ | Design matrix for fixed effects |
| $\beta$ | Vector of fixed effects (e.g., block, check) |
| $Z$ | Incidence matrix linking genotypes to plots |
| $g$ | Genetic effects, $g \sim N(0,\, \sigma_g^2 K)$ |
| $K$ | Genetic relationship matrix (GRM or A) |
| $e$ | Residuals, $e \sim N(0,\, R)$ |

The design problem is to choose $X$ and $Z$ — through field layout
construction — to optimise the precision of $\hat{\beta}$ (BLUEs) and the
prediction accuracy of $\hat{g}$ (BLUPs).

### Field as a spatial process

Field plots are spatially structured. Residuals are rarely independent owing
to row and column effects, soil gradients, and management variation. Three
residual structures are supported:

- **IID** — independent and identically distributed residuals (classical
  assumption)
- **AR1** — first-order autoregressive structure along one direction
- **AR1×AR1** — separable two-dimensional autoregressive structure (rows and
  columns)

### Genetic structure in design

Entries can be grouped based on family labels, the pedigree relationship
matrix $A$, or the genomic relationship matrix (GRM). Incorporating genetic
structure at the design stage allows the layout to reduce clustering of
related genotypes, improve diversity spread across blocks, and reduce
confounding between spatial and genetic signals.

---

## Design Strategies

### Block-based designs with repeated checks — `prep_famoptg()`

`prep_famoptg()` constructs a block-based design in which checks are repeated
in every block, while non-check entries are allocated across blocks according
to their specified replication levels. Depending on the replication pattern
supplied, the same function generates:

- an **augmented design** — unreplicated test entries with checks in every
  block
- a **partially replicated (p-rep) design** — some entries replicated, others
  unreplicated
- an **RCBD-type design** — all non-check entries replicated equally, with
  `n_blocks` equal to the replication number

The key design rule is that **a treatment may appear at most once per block**,
regardless of its total replication. This ensures replicates are distributed
across blocks rather than clustered.

**Use `prep_famoptg()` when:**

- checks must appear in every block
- many entries must be screened with limited resources
- the design may be augmented, p-rep, or RCBD-type depending on replication
- early- to intermediate-stage breeding evaluation is the objective

### Stream-based alpha designs — `alpha_rc_stream()`

`alpha_rc_stream()` constructs alpha row–column designs on a fixed grid using
a stream-based layout. The field is converted into a single ordered stream of
positions, split into contiguous replicate segments, and further divided into
incomplete blocks.

A key property of this approach is that **unused cells are placed only at the
end of the field stream** — not scattered across the grid — which is critical
for practical field implementation.

**Use `alpha_rc_stream()` when:**

- field dimensions are fixed and cannot change
- planting follows field-book (serpentine) order
- replicates are defined operationally as contiguous segments
- incomplete blocks may be unequal in size
- classical rectangular alpha designs are not feasible

---

## Example Dataset

The package ships with a built-in example dataset containing inputs for both
main functions:

```{r}
data("OptiDesign_example_data", package = "OptiDesign")
x <- OptiDesign_example_data
names(x)
```

---

## Example 1 — Family-based block design with repeated checks

This example calls `prep_famoptg()` with family-based grouping. Depending on
the replication pattern in the example inputs, the function acts as an
augmented, p-rep, or RCBD-type constructor:

```{r}
out_fam <- do.call(
  OptiDesign::prep_famoptg,
  c(
    x$OptiDesign_famoptg_example,
    x$OptiDesign_famoptg_args_family
  )
)

dim(out_fam$layout_matrix)
head(out_fam$field_book)
```

The result contains:

- `layout_matrix` — a spatial grid of plot assignments (`n_rows × n_cols`)
- `field_book` — a row-by-row record of plot, block, treatment, and grouping
  information suitable for field implementation

In this design: checks are present in every block; non-check entries are
allocated across blocks according to their replication levels; no non-check
entry appears twice in the same block; and grouping is based on family labels.

---

## Example 2 — Stream-based alpha row–column design

```{r}
out_alpha <- do.call(
  OptiDesign::alpha_rc_stream,
  c(
    x$OptiDesign_alpha_example,
    x$OptiDesign_alpha_args_family
  )
)

dim(out_alpha$layout_matrix)
head(out_alpha$field_book)
```

In this design: the field is traversed as a single ordered stream; replicate
segments are contiguous within that stream; checks are placed in every
incomplete block; and any unused cells appear only at the end of the stream.

---

## Example 3 — GRM-based stream design

When a genomic relationship matrix is available it can replace family labels
as the grouping source, allowing the layout to reflect genomic similarity more
precisely:

```{r}
out_grm <- do.call(
  OptiDesign::alpha_rc_stream,
  c(
    x$OptiDesign_alpha_example,
    x$OptiDesign_alpha_args_grm
  )
)

head(out_grm$field_book)
```

---

## Grouping Options

Both functions support three grouping strategies:

| Strategy | Source | Best for |
|---|---|---|
| Family-based | User-defined labels | Simple, interpretable grouping |
| GRM-based | Genomic similarity matrix | Captures real genomic relationships |
| Pedigree-based | $A$ matrix | When genomic data is unavailable |

---

## Dispersion Optimization

An optional step that improves the spatial distribution of genetically related
entries. If $d(i,j)$ denotes the similarity between plots $i$ and $j$, the
optimizer seeks to reduce:

$$\sum_{(i,j)\in \mathcal{N}} d(i,j)$$

where $\mathcal{N}$ is the set of neighbouring plot pairs. This reduces local
clustering of related materials, improves robustness to spatial confounding,
and promotes better distribution of genetic diversity across the field.

Use dispersion optimization when genomic or pedigree structure is important
and strong local spatial correlation is expected.

---

## Efficiency Evaluation

An optional diagnostic step that evaluates a candidate design under a
mixed-model framework before field implementation.

**Fixed-effect precision** — the variance of BLUE estimates:

$$\mathrm{Var}(\hat{\beta}) = (X^\top V^{-1} X)^{-1}$$

Lower diagonal values indicate higher precision for the corresponding fixed
effects.

**Random-effect prediction** — the prediction error variance (PEV):

$$\mathrm{PEV} = \mathrm{Var}(\hat{g} - g)$$

Lower PEV values indicate higher prediction accuracy for genetic effects.

Supported models include IID, AR1, and AR1×AR1 spatial structures, and
prediction modes include BLUP, GBLUP, and PBLUP. Use efficiency evaluation
when comparing alternative designs or studying the impact of block structure
and replication on estimability.

---

## Practical Guidelines

| Criterion | `prep_famoptg()` | `alpha_rc_stream()` |
|---|---|---|
| Checks in every block | Yes — by design | Yes — by design |
| Field dimensions | Flexible | Fixed grid required |
| Replicate structure | Block-based | Contiguous stream segments |
| Unequal block sizes | Not applicable | Supported |
| Unused cell placement | Depends on layout | End of stream only |
| Typical use | Early/mid breeding stage | Operational field-book trials |

**Grouping choice:** use family labels for simple interpretability, GRM when
genomic data is available and relatedness precision matters, and the $A$
matrix when only pedigree information is available.

---

## Summary

`OptiDesign` provides two complementary design strategies — block-based
repeated-check layouts via `prep_famoptg()` and stream-based alpha row–column
layouts via `alpha_rc_stream()` — together with optional dispersion
optimization and mixed-model efficiency evaluation. Both functions support
family, genomic, and pedigree-based grouping and are designed to be
practically implementable in field trials with real-world constraints.

---

## Session Information

```{r}
sessionInfo()
```
